### 2024-04-01 paper/paper.qmd
pdf has enough for peer review submission

---
title: "Homeless Shelter Usage Trends and Mortality: A Correlation Analysis"
author: 
  - Maroosh Gillani
thanks: "Code and data are available at: https://github.com/Maroosh-Gillani/homeless_mortality."
date: today
date-format: long
abstract: "This paper investigates the relationship between homeless shelter usage trends and mortality among those experiencing homelessness in Toronto. Using datasets from Open Data Toronto, I analyzed patterns of shelter occupancy and transitions alongside occurrences of death among homeless individuals on a yearly basis. The findings reveal a significant correlation between shelter use patterns such as returns to shelters, and an increased risk of mortality in those months. Understanding these correlations showcases the dynamics between shelter access, economic instability, and mortality risks faced by those experiencing homelessness in Toronto, hoping to illuminate these issues, raise awareness, and to prompt reflection on potential avenues for assisting these individuals."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

In Toronto, homelessness remains a persistent social issue with profound implications for public health and social welfare [@homelessNewsArticle]. Amidst efforts to address homelessness, understanding the intricate dynamics between shelter usage patterns and mortality risks among individuals experiencing homelessness is crucial. This paper delves into this relationship, aiming to shed light on the underlying factors that contribute to mortality within this vulnerable population.

Using data sets sourced from Open Data Toronto [@opendatatoronto], this study examines the interplay between homeless shelter usage trends and mortality among individuals experiencing homelessness in Toronto. By analyzing patterns of shelter occupancy, transitions, and occurrences of death on a yearly basis, I aim to highlight the connections between shelter access, economic instability, and mortality risks faced by those without stable housing. Additionally, I will employ a logistic regression model to further explore the relationships between shelter usage patterns and mortality outcomes, providing statistical insights into these dynamics.

The findings of this study reveal a notable correlation between specific shelter use patterns, such as returning to shelters, and an increased risk of mortality during certain months. This contributes to the understanding of challenges faced by individuals experiencing homelessness and emphasizes the importance of targeted interventions to address mortality risks within this population.

By discussing these issues and raising awareness of the interconnected factors influencing homelessness and mortality, this paper aims to encourage reflection on potential ways to support individuals experiencing homelessness in Toronto. Through this exploration, I aim to underscore the significance of addressing the underlying causes of homelessness and advocating for approaches to safeguard the well-being and dignity of all members of society.

# Data {#sec-data}

All data collection and analysis was done using statistical tools such as R [@citeR].

# Measurement

# Model

```{r}
#| include: false
#| warning: false
#| message: false

# consider using logistic regression -> usage patterns (independent variables) and mortality (dependent variable).
```

## Model set-up

### Model justification

# Results

# Discussion

\newpage

# References

### 2024-04-02 paper/paper.qmd
update abstract typo

@@
 The findings reveal a significant correlation between shelter use patterns such as returns to shelters, and an increased risk of mortality in those months.
-Understanding these correlations showcases the dynamics between shelter access, economic instability, and mortality risks faced by those experiencing homelessness in Toronto, hoping to illuminate these issues, raise awareness, and to prompt reflection on potential avenues for assisting these individuals."
+Understanding these correlations showcases the dynamics between shelter access, economic instability, and mortality risks faced by those experiencing homelessness in Toronto, and helps illuminate these issues, raise awareness, and to prompt reflection on potential avenues for assisting these individuals."
 format: pdf

### 2024-04-12 scripts/00-simulate_data.R
added simulation

@@ -1,19 +1,28 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Simulates number of homeless deaths and number of homeless people returning to shelters per year
+# Author: Maroosh Gillani
+# Date: 12 March 2024
+# Contact: maroosh.gillani@mail.utoronto.ca
+# Pre-requisites: N/A
+# Any other information needed? N/A
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(7)
+
+years <- 2018:2022
+
+returns <- seq(1000, 1500, length.out = length(years)) + rnorm(length(years), mean = 0, sd = 100)
+deaths <- returns * 0.02 + rnorm(length(years), mean = 0, sd = 20)
+
+# store simulated data in a dataframe
+yearly_data <- data.frame(Year = years,
+                   `Number of Homeless Deaths` = round(deaths),
+                   `Number of People Returning to Homeless Shelters` = round(returns))
+
 
 
 

### 2024-04-14 scripts/00-simulate_data.R
cleaned data + fixed date in simulate_data

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates number of homeless deaths and number of homeless people returning to shelters per year
 # Author: Maroosh Gillani
-# Date: 12 March 2024
+# Date: 12 April 2024
 # Contact: maroosh.gillani@mail.utoronto.ca
 # Pre-requisites: N/A
 # Any other information needed? N/A

### 2024-04-14 scripts/02-data_cleaning.R

@@ -1,44 +1,54 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Cleans the raw homeless death and Toronto shelter data originally from OpenDataToronto
+# Author: Maroosh Gillani
+# Date: 13 April 2024
+# Contact: maroosh.gillani@mail.utoronto.ca
+# Pre-requisites: make sure you have successfully ran 01-download_data.R
+# Any other information needed? N/A
 
 #### Workspace setup ####
 library(tidyverse)
+library(dplyr)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+raw_death_data <- read_csv("data/raw_data/raw_homeless_death_data.csv")
+raw_shelter_data <- read.csv("data/raw_data/raw_shelter_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+# Clean raw_death_data
+cleaned_death_data <- raw_death_data %>%
[…]

### 2024-04-15 scripts/02-data_cleaning.R
updated clean data to include months as well

@@ -14,41 +14,54 @@ library(dplyr)
 raw_death_data <- read_csv("data/raw_data/raw_homeless_death_data.csv")
 raw_shelter_data <- read.csv("data/raw_data/raw_shelter_data.csv")
 
-# Clean raw_death_data
+
+month_mapping <- c("January" = 1, "February" = 2, "March" = 3, "April" = 4, "May" = 5, "June" = 6, "July" = 7, "August" = 8, "September" = 9, "October" = 10, "November" = 11, "December" = 12)
+
+# Cleaned death data with year and month
 cleaned_death_data <- raw_death_data %>%
-  # Select only the "Year of death" and "Count" columns
-  select(`Year of death`, Count) %>%
+  # Select only the "Year of death", "Month of death", and "Count" columns
+  select(`Year of death`, `Month of death`, Count) %>%
   # Filter rows where the year is between 2018 and 2022
   filter(`Year of death` >= 2018 & `Year of death` <= 2022) %>%
-  # Group by "Year of death" and summarize the counts
-  group_by(`Year of death`) %>%
-  summarize(Count = sum(Count))
+  # Group by "Year of death" and "Month of death" and summarize the counts
+  group_by(`Year of death`, `Month of death`) %>%
+  # summarize(Count = sum(Count)) %>%
+  # Convert month names to numeric values using custom mapping
+  mutate(Month = month_mapping[`Month of death`])
 
-# Clean raw_shelter_data
+
+# Cleaned shelter data with year and month
 cleaned_shelter_data <- raw_shelter_data %>%
-  # Select only the relevant columns
+  # Select only the "date.mmm.yy.", "population_group", and "returned_to_shelter" columns
   select(date.mmm.yy., population_group, returned_to_shelter) %>%
-  # Filter rows where the year is between 18 and 22
+  # Filter rows where the year is between 2018 and 2022
   filter(as.integer(substring(date.mmm.yy., nchar(date.mmm.yy.) - 1)) >= 18 & as.integer(substring(date.mmm.yy., nchar(date.mmm.yy.) - 1)) <= 22) %>%
   # Filter rows where the population_group is "All Population"
   filter(population_group == "All Population") %>%
-  # Group by year and summarize the counts of "returned_to_shelter"
-  group_by(year = as.integer(substring(date.mmm.yy., nchar(date.mmm.yy.) - 1))) %>%
[…]

### 2024-04-15 scripts/02-data_cleaning.R
improved column names for clean data

@@ -19,31 +19,19 @@ month_mapping <- c("January" = 1, "February" = 2, "March" = 3, "April" = 4, "May
 
 # Cleaned death data with year and month
 cleaned_death_data <- raw_death_data %>%
-  # Select only the "Year of death", "Month of death", and "Count" columns
   select(`Year of death`, `Month of death`, Count) %>%
-  # Filter rows where the year is between 2018 and 2022
   filter(`Year of death` >= 2018 & `Year of death` <= 2022) %>%
-  # Group by "Year of death" and "Month of death" and summarize the counts
   group_by(`Year of death`, `Month of death`) %>%
-  # summarize(Count = sum(Count)) %>%
-  # Convert month names to numeric values using custom mapping
   mutate(Month = month_mapping[`Month of death`])
 
-
-# Cleaned shelter data with year and month
+# Cleaned shelter data
 cleaned_shelter_data <- raw_shelter_data %>%
-  # Select only the "date.mmm.yy.", "population_group", and "returned_to_shelter" columns
   select(date.mmm.yy., population_group, returned_to_shelter) %>%
-  # Filter rows where the year is between 2018 and 2022
   filter(as.integer(substring(date.mmm.yy., nchar(date.mmm.yy.) - 1)) >= 18 & as.integer(substring(date.mmm.yy., nchar(date.mmm.yy.) - 1)) <= 22) %>%
-  # Filter rows where the population_group is "All Population"
   filter(population_group == "All Population") %>%
-  # Extract year and month from "date.mmm.yy." column
   mutate(Year = paste0("20", substring(date.mmm.yy., nchar(date.mmm.yy.) - 1)),
          Month = match(substring(date.mmm.yy., 1, 3), month.abb)) %>%
-  # Select only the "Year", "Month", and "returned_to_shelter" columns
   select(Year, Month, returned_to_shelter) %>%
-  # Group by "Year" and "Month" and summarize the counts
   group_by(Year, Month) %>%
   summarize(returned_to_shelter = sum(returned_to_shelter))
 
@@ -62,6 +50,11 @@ clean_data <- clean_data %>%
 # Remove the separate Month and Year columns
 clean_data <- clean_data %>%
   select(-Year, -Month)
+
+# Rename columns
+clean_data <- clean_data %>%
+  rename(`Returned to shelter` = returned_to_shelter,
+         `Month/Year` = Year_Month)
 
 #### Save data ####
 write_csv(clean_data, "data/clean_data/cleaned_data.csv")

### 2024-04-15 scripts/02-data_cleaning.R
change column name again

@@ -54,7 +54,7 @@ clean_data <- clean_data %>%
 # Rename columns
 clean_data <- clean_data %>%
   rename(`Returned to shelter` = returned_to_shelter,
-         `Month/Year` = Year_Month)
+         `Month/Year` = Year_Month, `Deaths` = Count)
 
 #### Save data ####
 write_csv(clean_data, "data/clean_data/cleaned_data.csv")

### 2024-04-15 scripts/03-test_data.R
added tests for cleaned data

@@ -1,15 +1,35 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests cleaned data from OpenDataToronto's homeless deaths and homeless shelter datasets
+# Author: Maroosh Gillani
+# Date: 15 April 2024
+# Contact: maroosh.gillani@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Make sure that 01-download_data.R and 02-data_cleaning.R have been run already.
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(here)
+library(haven)
+
+# read data
+cleaned_data = read.csv(file=here("data/clean_data/cleaned_data.csv"))
 
 #### Test data ####
+
+# Returned to shelter and Deaths are numeric and non-negative
+all(cleaned_data$Returned.to.shelter >= 0) & all(cleaned_data$Deaths >= 0)
+
+# total of 60 rows (12 months x 5 years)
+nrow(cleaned_data) == 60
+
+# months are in between 1 and 12
+all(sapply(strsplit(cleaned_data$Month.Year, "/"), 
+           function(x) as.integer(x[1]) >= 1 & as.integer(x[1]) <= 12))
+
+# no missing values
+!any(is.na(cleaned_data$Returned.to.shelter)) & !any(is.na(cleaned_data$Deaths)) &
+  !any(is.na(cleaned_data$Month.Year))
+
+# all unique rows
+nrow(unique(cleaned_data)) == nrow(cleaned_data)

### 2024-04-15 paper/paper.qmd
added to model script + model rds file

@@
 library(palmerpenguins)
+
+# THIS IS TEMPORARILY HERE -- POISSON MODEL GRAPH
+predicted_counts <- predict(poisson_model, type = "response")
+
+# Create a scatter plot of observed vs. predicted counts
+plot(cleaned_data$Returned.to.shelter, cleaned_data$Deaths, 
+     xlab = "Returned to shelter", ylab = "Deaths",
+     main = "Observed vs. Predicted Counts",
+     pch = 16, col = "blue")
+lines(cleaned_data$Returned.to.shelter, predicted_counts, col = "red", type = "l", lwd = 2)
+legend("topright", legend = c("Observed", "Predicted"), 
+       col = c("blue", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))
+
 ```

### 2024-04-15 scripts/04-model.R

@@ -1,37 +1,27 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Models the cleaned data derived from OpenDataToronto's datasets using poisson regression
+# Author: Maroosh Gillani
+# Date: 15 April 2024
+# Contact: maroosh.gillani@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: Make sure that 01-download_data.R and 02-data_cleaning.R have been run already.
 
 #### Workspace setup ####
 library(tidyverse)
+library(here)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+cleaned_data = read.csv(file=here("data/clean_data/cleaned_data.csv"))
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
-
+poisson_model <- 
+  glm(Deaths ~ Returned.to.shelter, data = cleaned_data, family = poisson)
 
 #### Save model ####
 saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+  poisson_model,
+  file = "models/poisson_model.rds"
 )
 
 

### 2024-04-15 paper/paper.qmd
made some graphs

--- 2024-04-02 paper/paper.qmd
@@
 
+# load libraries
 library(tidyverse)
-library(palmerpenguins)
+library(here)
+library(ggplot2)
+
+# read in cleaned data
+cleaned_data = read.csv(file=here("data/clean_data/cleaned_data.csv"))
 ```
@@
 
-# consider using logistic regression -> usage patterns (independent variables) and mortality (dependent variable).
+# THIS IS TEMPORARILY HERE -- POISSON MODEL GRAPH
+#predicted_counts <- predict(poisson_model, type = "response")
+
+# Create a scatter plot of observed vs. predicted counts
+#plot(cleaned_data$Returned.to.shelter, cleaned_data$Deaths, 
+#     xlab = "Returned to shelter", ylab = "Deaths",
+#     main = "Observed vs. Predicted Counts",
+#     pch = 16, col = "blue")
+#lines(cleaned_data$Returned.to.shelter, predicted_counts, col = "red", type = "l", lwd = 2)
+#legend("topright", legend = c("Observed", "Predicted"), 
+#       col = c("blue", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = #c(NA, 2))
 ```
@@
 # Results
+```{r graph1, fig.cap="temp"}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-1
+
+# graph monthly homeless deaths over 2018 - 2022
+
+# Convert Month.Year to factor to preserve the order of months
+cleaned_data$Month.Year <- factor(cleaned_data$Month.Year, levels = unique(cleaned_data$Month.Year))
+
+ggplot(cleaned_data, aes(x = Month.Year, y = Deaths)) +
+  geom_line(aes(group = 1), color = "red", size = 1) +
+  labs(x = "Month/Year", y = "Deaths", title = "Monthly Homeless Deaths") +
+  scale_x_discrete(breaks = cleaned_data$Month.Year[seq(1, length(cleaned_data$Month.Year), by = 6)]) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+@fig-1 says this that and whatever.
+
+```{r graph2, fig.cap="temp2"}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-2
+
+# graph monthly returns to shelters over 2018 - 2022
+
+# Convert Month.Year to factor to preserve the order of months
+cleaned_data$Month.Year <- factor(cleaned_data$Month.Year, levels = unique(cleaned_data$Month.Year))
+
+ggplot(cleaned_data, aes(x = Month.Year, y = Returned.to.shelter)) +
+  geom_line(aes(group = 1), color = "blue", size = 1) +
+  labs(x = "Month/Year", y = "Returns to Homeless Shelters", title = "Monthly Returns to Homeless Shelters") +
+  scale_x_discrete(breaks = cleaned_data$Month.Year[seq(1, length(cleaned_data$Month.Year), by = 6)]) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+@fig-2 says some stuff, other stuff and also stuff
 

### 2024-04-16 scripts/02-data_cleaning.R
added parquet

@@ -9,6 +9,7 @@
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
+library(arrow)
 
 #### Clean data ####
 raw_death_data <- read_csv("data/raw_data/raw_homeless_death_data.csv")
@@ -58,3 +59,5 @@ clean_data <- clean_data %>%
 
 #### Save data ####
 write_csv(clean_data, "data/clean_data/cleaned_data.csv")
+write_parquet(clean_data, "data/clean_data/cleaned_data.parquet")
+

### 2024-04-17 paper/paper.qmd
updated citations for packages

@@
 library(ggplot2)
+library(knitr)
 
@@
 
-In Toronto, homelessness remains a persistent social issue with profound implications for public health and social welfare [@homelessNewsArticle].
-Amidst efforts to address homelessness, understanding the intricate dynamics between shelter usage patterns and mortality risks among individuals experiencing homelessness is crucial.
-This paper delves into this relationship, aiming to shed light on the underlying factors that contribute to mortality within this vulnerable population.
+In Toronto, homelessness remains a persistent social issue with significant implications for public health and social welfare [@homelessNewsArticle].
+Amidst efforts to address homelessness, understanding the dynamics between shelter usage patterns and mortality risks among individuals experiencing homelessness is crucial.
+This paper looks into this relationship, aiming to analyze the underlying factors that contribute to mortality within this vulnerable population.
 
-Using data sets sourced from Open Data Toronto [@opendatatoronto], this study examines the interplay between homeless shelter usage trends and mortality among individuals experiencing homelessness in Toronto.
-By analyzing patterns of shelter occupancy, transitions, and occurrences of death on a yearly basis, I aim to highlight the connections between shelter access, economic instability, and mortality risks faced by those without stable housing.
-Additionally, I will employ a logistic regression model to further explore the relationships between shelter usage patterns and mortality outcomes, providing statistical insights into these dynamics.
+Using data sets sourced from Open Data Toronto [@opendatatoronto], this study examines the interplay between homeless shelter usage trends, particularly returns to homeless shelters per month, and monthly mortality among individuals experiencing homelessness in Toronto.
+By analyzing patterns of shelter occupancy, transitions, and occurrences of death on a monthly basis, I aim to highlight the connections between shelter access, economic instability, and mortality risks faced by those without stable housing.
+Additionally, I will employ a poisson regression model to further explore the relationships between shelter return patterns and mortality outcomes, providing statistical insights into this dynamic.
 
-The findings of this study reveal a notable correlation between specific shelter use patterns, such as returning to shelters, and an increased risk of mortality during certain months.
+The findings of this study reveal a notable correlation between the pattern of returning to shelters, and an increased risk of mortality during certain months.
 This contributes to the understanding of challenges faced by individuals experiencing homelessness and emphasizes the importance of targeted interventions to address mortality risks within this population.
@@
 
-All data collection and analysis was done using statistical tools such as R [@citeR].
+This paper uses data sets from OpenDataToronto [@opendatatoronto], which provides access to freely available digital data from the city of Toronto.
+In particular, the "Deaths of People Experiencing Homelessness" and "Toronto Shelter System Flow" data sets are being utilized for the purposes of analyzing the number of people returning to homeless shelters, and the number of homeless people who passed away over the time period of 2018 to 2022.
+
+All data collection, analysis, modeling and visualization was done using R [@citeR].
+In particular, the following packages were utilized: tidyverse [@tidyverse], ggplot2 [@citeggplot2], here [@citehere], knitr [@citeknitr], arrow [@citearrow], dplyr [@citedplyr], and haven [@citehaven].
 
@@
 # Model
+
+MAKE A NULL HYPOTHESIS AND ALTERNATE HYPOTHESIS FOR POISSON REGRESSION
 
@@
 ```
-@fig-1 says this that and whatever.
+@fig-1 shows that deaths among homeless people in Toronto, while fluctuating significantly over the months, are following a general trend of growth.
+Each year, the deaths have a consistent peak around winter time, especially near the month of January.
+The highest number of homeless deaths in Toronto seem to have occurred in the November of 2021, while the lowest amount of deaths recorded are in December of 2018.
 
@@
 
-@fig-2 says some stuff, other stuff and also stuff
+@fig-2, while also fluctuating significantly, seems to follow a negative pattern for returns of homeless people to shelters.
+The returns are at their peak in November of 2019.
+In contrast, September of 2022 saw the least amount of Toronto's homeless population return to a shelter.
+
+```{r graph3, fig.cap="temp3"}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-3
+
+poisson_model <- readRDS(here("models/poisson_model.rds"))
+
+model_summary <- summary(poisson_model)
+
+model_summary_df <- as.data.frame(model_summary$coefficients)
+
+model_table <- knitr::kable(model_summary_df)
+
+model_table
+```
+@fig-3 shows the results from the Poisson regression conducted on the cleaned dataset.
+The aim of this model was to investigate the relationship between the number of homeless deaths and the number of individuals returning to homeless shelters.
+In particular, "Returned to shelter" was used as the sole predictor variable for monthly homeless deaths.
+
+Firstly, the intercept represents the estimated log count of homeless deaths when the predictor variable is 0.
+Here, we can see that it is 3.0368 (with SE = 0.1828, z = 16.6157 and p = 0), which represents the base level of homeless deaths when no individuals are returning to shelters.
+
+The estimated coefficient for Returned to shelter is approximately -0.0011 (with SE = 0.0004, z = -2.6511, and p = 0.0080).
+This represents the estimated change in the log count of homeless deaths in relation with a one unit increase in the number of individuals returning to homeless shelters per month.
+Since the coefficient is negative, the average increase of people returning to homeless shelters can be associated with a decrease in the number of homeless death.
 

### 2024-04-17 scripts/04-model.R

@@ -9,7 +9,6 @@
 #### Workspace setup ####
 library(tidyverse)
 library(here)
-library(rstanarm)
 
 #### Read data ####
 cleaned_data = read.csv(file=here("data/clean_data/cleaned_data.csv"))

### 2024-04-18 paper/paper.qmd
updated figure captions, added to models and measurement sections, started working on discussion, updated references

@@
 
-All data collection, analysis, modeling and visualization was done using R [@citeR].
+The "Deaths of People Experiencing Homelessness" is very simple, and consists of the year and months (starting January 2018), a unique id for each month, and the count of homeless persons who passed away in that month in Toronto.
+All columns in this dataset are numeric, with the exception of 'Month of death', which consists of characters.
+For this analysis, I have utilized the columns of 'Year of death', 'Month of death', and 'Count' only.
+
+The "Toronto Shelter System Flow" is a more complex dataset, consisting of many columns such as 'date(mmm-yy)', and 'population_group', which contain characters, and details regarding the population relevant to homeless shelters in Toronto, such as 'returned_from_housing' or 'became_inactive', which are numeric in nature.
+
+There were no similar data sets to my knowledge.
+Furthermore, the two data sets used in this paper are ideal as they are coming directly from the city of Toronto, and the organizations and agencies dealing with homeless shelters and recording the deaths of homeless individuals.
+As such, alternate data sets do not need to be considered.
+
+The data sets were cleaned to reformat the date into 'mm/yyyy' format.
+This allowed for me to extract and merge the relevant information from both data sets easily into one table.
+This aided in utilizing and having access to the necessary information only when making graphs and models for analysis purposes.
+
+The most important variables are 'Month/Year', 'Deaths', and 'Returns to Homeless Shelters'.
+'Month/Year' is the independent variable for both @fig-1 and @fig-2.
+'Deaths' is the dependent variable for @fig-1, while 'Returns to Homeless Shelters' is the dependent variable for @fig-2.
+All of these variables were also key for the Poisson regression model utilized in this analysis, and more details can be seen in the Model section.
+
+All data collection, analysis, modeling and visualization was done using R [@citeR], with the aid of the RStudio IDE [@citeRStudio] to streamline the process.
 In particular, the following packages were utilized: tidyverse [@tidyverse], ggplot2 [@citeggplot2], here [@citehere], knitr [@citeknitr], arrow [@citearrow], dplyr [@citedplyr], and haven [@citehaven].
@@
 
+The "Deaths of People Experiencing Homelessness" data set is maintained by the Toronto Public Health, particularly in regards to data collection, reporting and analysis [@opendatatoronto].
+It is refreshed semi-annually.
+Their information is supplied to them by agencies such as the SSHA, alongside other social service agencies relevant to homelessness.
+
+On the other hand, the "Toronto Shelter System Flow" gets its data from an information management system that is used to operate shelter related services that are funded by the City of Toronto, and it is updated on a monthly basis [@opendatatoronto].
+
 # Model
 
-MAKE A NULL HYPOTHESIS AND ALTERNATE HYPOTHESIS FOR POISSON REGRESSION
+## Model justification
 
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-# THIS IS TEMPORARILY HERE -- POISSON MODEL GRAPH
-#predicted_counts <- predict(poisson_model, type = "response")
-
-# Create a scatter plot of observed vs. predicted counts
-#plot(cleaned_data$Returned.to.shelter, cleaned_data$Deaths, 
-#     xlab = "Returned to shelter", ylab = "Deaths",
-#     main = "Observed vs. Predicted Counts",
-#     pch = 16, col = "blue")
-#lines(cleaned_data$Returned.to.shelter, predicted_counts, col = "red", type = "l", lwd = 2)
-#legend("topright", legend = c("Observed", "Predicted"), 
-#       col = c("blue", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = #c(NA, 2))
-```
+To model the relationship between the number of homeless deaths and the number of homeless individuals returning to shelters on a monthly basis, a Poisson regression model was utilized.
+Such a model is suitable for this purpose because the number of deaths, as well as the number of people returning to shelters, are both instances of count data.
+The outcome variable here represents the counts of events occurring within a fixed interval.
 
@@
 
-### Model justification
+The model is specified as follows:
+$$log(Expected Number of Homeless Deaths) = \beta_{0} + \beta{1}(Number of Individuals Returning to Shelters)$$
+Here, $\beta_{0}$ represents the intercept, or the expected log count of homeless deaths when the number of people returning to shelters is 0.
+
+$\beta_{1}$ represents the coefficient for the predictor variable, indicating the change in the expected log count of homeless deaths for each unit increase in the number of individuals returning to homeless shelters.
 
 # Results
-```{r graph1, fig.cap="temp"}
+```{r graph1, fig.cap="Deaths of Homeless Individuals in Toronto per month (2018 - 2022)"}
 #| echo: false
@@
   geom_line(aes(group = 1), color = "red", size = 1) +
-  labs(x = "Month/Year", y = "Deaths", title = "Monthly Homeless Deaths") +
+  labs(x = "Month/Year", y = "Deaths") +
   scale_x_discrete(breaks = cleaned_data$Month.Year[seq(1, length(cleaned_data$Month.Year), by = 6)]) +
@@
 
-```{r graph2, fig.cap="temp2"}
+```{r graph2, fig.cap="Number of Individuals Returning to Homeless Shelters per Month (2018 - 2022)"}
 #| echo: false
@@
   geom_line(aes(group = 1), color = "blue", size = 1) +
-  labs(x = "Month/Year", y = "Returns to Homeless Shelters", title = "Monthly Returns to Homeless Shelters") +
+  labs(x = "Month/Year", y = "Returns to Homeless Shelters") +
   scale_x_discrete(breaks = cleaned_data$Month.Year[seq(1, length(cleaned_data$Month.Year), by = 6)]) +
@@
 
-```{r graph3, fig.cap="temp3"}
+```{r graph3, fig.cap="Summary of the Results from The Poisson Regression Model"}
 #| echo: false
@@
 
+Looking at both @fig-1 and @fig-2, it can be seen that both graphs are growing in opposite directions over the same period of time.
+Homeless deaths seem to be rising, while the amount of homeless people returning to shelters seems to be in decline.
+This implies that when more homeless people returned to shelters, the mortality within this population was a lot lower.
+This is further supported by the Poisson regression model run on this same data, which posits that for each homeless person that returns to a shelter, the average number of deaths in the homeless population decreases, as seen in @fig-3.
+
+Thus, data suggests that homeless shelters are indeed beneficial towards the homeless population of Toronto from a mortality standpoint.
+Then, why is that as time passes, returns to homeless shelters decrease?
+Perhaps the issue is not as straightforward as it initially seems.
+@shelterdeathtoronto highlights that both 2021 and 2022 consecutively saw over 100 deaths among homeless shelter users in Toronto.
+This implies that homeless people in Toronto are at risk of death regardless of having access to shelter services.
+Another implication could be that the services and resources provided by these shelters are inadequate to ensure homeless survival, which could be why the homeless population in Toronto has progressively chosen to return to shelters less with the passage of
+
+Limitations: cause of death not known in data sets.
+
 \newpage

### 2024-04-18 paper/paper.qmd
finish paper, update references

@@
 ---
-title: "Homeless Shelter Usage Trends and Mortality: A Correlation Analysis"
+title: "Return to Homeless Shelters and Mortality: A Correlation Analysis"
 author: 
@@
 date-format: long
-abstract: "This paper investigates the relationship between homeless shelter usage trends and mortality among those experiencing homelessness in Toronto.
-Using datasets from Open Data Toronto, I analyzed patterns of shelter occupancy and transitions alongside occurrences of death among homeless individuals on a yearly basis.
-The findings reveal a significant correlation between shelter use patterns such as returns to shelters, and an increased risk of mortality in those months.
-Understanding these correlations showcases the dynamics between shelter access, economic instability, and mortality risks faced by those experiencing homelessness in Toronto, and helps illuminate these issues, raise awareness, and to prompt reflection on potential avenues for assisting these individuals."
+abstract: "This paper investigates the relationship between returns to homeless shelters and mortality among those experiencing homelessness in Toronto.
+Using datasets from Open Data Toronto, I analyzed patterns of returns to shelters alongside occurrences of death among homeless individuals on a monthly basis.
+The findings reveal a correlation between returns to shelter patterns, and an decreased risk of mortality in those months.
+Understanding these correlations showcases the dynamics between shelter access, flaws in shelter operation, and mortality risks faced by those experiencing homelessness in Toronto, and helps illuminate these issues, raise awareness, and to prompt reflection on potential avenues for assisting these individuals."
 format: pdf
@@
 Additionally, I will employ a poisson regression model to further explore the relationships between shelter return patterns and mortality outcomes, providing statistical insights into this dynamic.
+The estimand of interest in this analysis is the effect of returns to homeless shelters per month on the monthly mortality rate among individuals experiencing homelessness.
+By estimating this parameter, I aim to quantify the impact of shelter usage patterns on mortality risks within this vulnerable population.
 
-The findings of this study reveal a notable correlation between the pattern of returning to shelters, and an increased risk of mortality during certain months.
+The findings of this study reveal a notable correlation between the pattern of returning to shelters, and a decreased risk of mortality during certain months.
 This contributes to the understanding of challenges faced by individuals experiencing homelessness and emphasizes the importance of targeted interventions to address mortality risks within this population.
@@
 Through this exploration, I aim to underscore the significance of addressing the underlying causes of homelessness and advocating for approaches to safeguard the well-being and dignity of all members of society.
+
+The structure of the paper is as follows: Sections 2 and 3 covers details surrounding the utilized data, and how it was measured.
+Section 4 expands upon the choice of model for this analysis.
+Section 5 visualizes the data, and finally, Section 6 discusses the implications surrounding the visualizations, and the broader homelessness issue in Toronto.
 
@@
 Firstly, the intercept represents the estimated log count of homeless deaths when the predictor variable is 0.
-Here, we can see that it is 3.0368 (with SE = 0.1828, z = 16.6157 and p = 0), which represents the base level of homeless deaths when no individuals are returning to shelters.
+Here, it can be seen that it is 3.0368 (with SE = 0.1828, z = 16.6157 and p = 0), which represents the base level of homeless deaths when no individuals are returning to shelters.
 
@@
 Since the coefficient is negative, the average increase of people returning to homeless shelters can be associated with a decrease in the number of homeless death.
+\newpage
 
@@
 
-Looking at both @fig-1 and @fig-2, it can be seen that both graphs are growing in opposite directions over the same period of time.
-Homeless deaths seem to be rising, while the amount of homeless people returning to shelters seems to be in decline.
-This implies that when more homeless people returned to shelters, the mortality within this population was a lot lower.
-This is further supported by the Poisson regression model run on this same data, which posits that for each homeless person that returns to a shelter, the average number of deaths in the homeless population decreases, as seen in @fig-3.
+Upon examining both @fig-1 and @fig-2, it is apparent that both graphs are moving in opposite directions over the same period of time.
+The data shows that homeless deaths seem to be rising, while the amount of homeless people returning to shelters seems to be in decline.
+This implies an inverse relationship between the two variables, indicating that when more homeless people returned to shelters, the mortality within this population was a lot lower.
+This observation is further supported by the Poisson regression model run on this same data, which posits that for each homeless person that returns to a shelter, the average number of deaths in the homeless population decreases, as illustrated in @fig-3.
 
-Thus, data suggests that homeless shelters are indeed beneficial towards the homeless population of Toronto from a mortality standpoint.
-Then, why is that as time passes, returns to homeless shelters decrease?
-Perhaps the issue is not as straightforward as it initially seems.
+These findings suggest that homeless shelters indeed play a role  in reducing mortality in the homeless population of Toronto.
+However, the declining trend of returns to shelters prompts the question of why this might be the case.
+A deeper look shows that the issue is not as straightforward as it initially seems.
 @shelterdeathtoronto highlights that both 2021 and 2022 consecutively saw over 100 deaths among homeless shelter users in Toronto.
 This implies that homeless people in Toronto are at risk of death regardless of having access to shelter services.
-Another implication could be that the services and resources provided by these shelters are inadequate to ensure homeless survival, which could be why the homeless population in Toronto has progressively chosen to return to shelters less with the passage of
+Another implication could be that the services and resources provided by these shelters are inadequate to ensure homeless survival, which could be why the homeless population in Toronto has progressively chosen to return to shelters less with the passage of time.
 
-Limitations: cause of death not known in data sets.
+This is further supported by @badconditions, who discusses the inhumane practices of Toronto's shelters.
+Hundreds of shelter seekers were forced to sleep on sidewalks in front of Toronto's referral center.
+In fact, in place of the city, frontline workers took the mantle of providing relief through necessities such as food, water and blankets to these individuals.
+In addition, church groups had to provide shelter to many of the shelter seekers in places of worship.
+Thus, it seems like the city of Toronto and its shelters are either ill equipped, or refuse to handle large number of homeless shelter seekers.
+This implies that the declining returns to shelters is not solely a consequence of refusal to access shelters on the homeless populations part, but rather a shortcoming of the current infrastructure in place, which is likely unable to support the demand.
+
+@badconditions also brings up more unethical practices such as multiple shelters having a lack of doors on washroom and shower stalls, withholding information regarding disease outbreaks in shelters, and refusing to meet basic needs in favor cost cutting, such as replacing bath towels with paper towels.
+All of these factors could also contribute to making shelters less appealing for homeless people, which could also be a factor impacting why returns to shelters has seen a negative trend in Toronto.
+
+So while it seems that homeless shelters do have a positive impact on the mortality rates among homeless populations in Toronto, it comes at the cost of enduring unpleasant conditions, including scenarios that bring up the trade-off between survival and dignity.
+In addition, even if homeless people want access to shelters, many are unable to access them due to poor infrastructure in referring these individuals to appropriate centers.
+This means that at the very least, significant reform is needed at a fundamental level for how homeless shelters are operated in Toronto.
+Some obvious changes include improving the shelter referral system such that it does not allow hundreds of individuals to be forced to live on sidewalks for extended durations.
+Another change is to improve conditions and treatment of homeless people within shelters.
+@fig-3 is clear in showing that homeless shelters reduce mortality, and making shelters more appealing to utilize can help in a significant way.
+
+A more ambitious solution would be to address the homelessness issue itself, and work towards eliminating homelessness at its root.
+@solution proposes multiple ways to combat homelessness in Canada, which include ideas such as the government implementing caps on rapidly increasing rent prices, fixing the National Housing Strategy by elevating the amount of affordable housing units being built, or by providing additional rent support to people at increased risks of becoming homeless.
+
+It must be noted that a significant limitation in this analysis is that the cause of death is not provided in the utilized data sets.
+Thus, it is a difficult to say how many deaths within homeless individuals can directly be affected by resources available at homeless shelters.
+Having access to more information surrounding causes of death, and other details such as location of death can help in providing a more robust analysis on factors contributing to Toronto's homeless mortality in relation to shelter usage.
 

### 2024-04-18 other/datasheet/datasheet.qmd
update readme + llm uage

---
title: "Datasheet for 'Deaths of People Experiencing Homelessness' and 'Toronto Shelter System Flow'"
author: 
  - Maroosh Gillani
thanks: "Code and data are available at: https://github.com/Maroosh-Gillani/homeless_mortality"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to enable analysis of Australian politicians. We were unable to find a publicly available dataset in a structured format that had the biographical and political information on Australian politicians that was needed for modelling.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - TBD
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - TBD
4. *Any other comments?*
    - TBD

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- TBD
2. *How many instances are there in total (of each type, if appropriate)?*
	- TBD
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- TBD
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- TBD
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- TBD
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- TBD
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- TBD
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- TBD
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- TBD
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- TBD
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- TBD
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- TBD
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- TBD
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- TBD
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- TBD
16. *Any other comments?*
	- TBD

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- TBD
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- TBD
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- TBD
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- TBD
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- TBD
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- TBD
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- TBD
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- TBD
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- TBD
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
12. *Any other comments?*
	- TBD

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- TBD
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- TBD
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- TBD
4. *Any other comments?*
	- TBD

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- TBD
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- TBD
3. *What (other) tasks could the dataset be used for?*
	- TBD
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- TBD
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- TBD
6. *Any other comments?*
	- TBD

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- TBD
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- TBD
3. *When will the dataset be distributed?*
	- TBD
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- TBD
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- TBD
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- TBD
7. *Any other comments?*
	- TBD

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- TBD
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- TBD
3. *Is there an erratum? If so, please provide a link or other access point.*
	- TBD
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- TBD
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- TBD
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- TBD
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- TBD
8. *Any other comments?*
	- TBD

\newpage

# References

### 2024-04-18 other/datasheet/datasheet.qmd
finished datasheet

@@
 1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
-    - The dataset was created to enable analysis of Australian politicians.
-We were unable to find a publicly available dataset in a structured format that had the biographical and political information on Australian politicians that was needed for modelling.
+    - The 'Deaths of People Experiencing Homelessness' dataset aims to track the deaths of Toronto's homeless population.
+    - The 'Toronto Shelter System Flow' dataset is used to track the details regarding Toronto's homeless shelters and their users.
 2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
-    - TBD
+    - The 'Deaths of People Experiencing Homelessness' dataset was created by Toronto Public Health (TPH) in collaboration with The Shelter, Support and Housing Administration (SSHA) and health and social service agencies.
+    - The 'Toronto Shelter System Flow' dataset was published by Toronto Shelter & Support Services, and the Shelter Management Information System (SMIS) is used to track the people using shelter services.
 3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
-    - TBD
+    - 'Deaths of People Experiencing Homelessness': not funded.
+    - 'Toronto Shelter System Flow': Funded by the City of Toronto.
 4. *Any other comments?*
-    - TBD
+    - No.
 
@@
 Please provide a description.*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': Each row shows the number of homeless people who passed away each month, starting from January 2017 to June 2023. 
+	- 'Toronto Shelter System Flow': 7 rows are dedicated to a single month (starting from January 2018).
+Of these seven rows, each row is dedicated to specific types of the population that is using Toronto's homeless shelters, such as Families, Youth, Refugees etc.
+For each different type of population, a count for how many fall into specific categories (such as newly identified, returned from housing, return to shelter, age groups, gender etc.) is provided.
 2. *How many instances are there in total (of each type, if appropriate)?*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': There are 81 rows, and each row has 4 columns. All columns are numeric except for Month of death, which is made of characters.
+	- 'Toronto Shelter System Flow': There are 556 rows, and each row has 18 columns. All columns are numeric except for date(mmm-yy) and population_group, which are characters.
 3.
@@
 If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
-	- TBD
+  - 'Deaths of People Experiencing Homelessness': The data is complete and contains all instances.
+	- 'Toronto Shelter System Flow': The data is complete and contains all instances.
 4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': Each row shows the number of homeless people who passed away each month, starting from January 2017 to June 2023. 
+	- 'Toronto Shelter System Flow': 7 rows are dedicated to a single month (starting from January 2018).
+Of these seven rows, each row is dedicated to specific types of the population that is using Toronto's homeless shelters, such as Families, Youth, Refugees etc.
+For each different type of population, a count for how many fall into specific categories (such as newly identified, returned from housing, return to shelter, age groups, gender etc.) is provided.
 5. *Is there a label or target associated with each instance? If so, please provide a description.*
-	- TBD
+	- Yes, both datasets have ids for each row, stored in a designated id column.
 6.
@@
 This does not include intentionally removed information, but might include, for example, redacted text.*
-	- TBD
+		- 'Deaths of People Experiencing Homelessness': The data only contains deaths of homeless people that are reported to the TPH by the SSHA partners.
+The data is also subject to change based on additional reports to TPH.
+	- 'Toronto Shelter System Flow': The dataset contains data from people who have used overnight services only, and is subject to change as new data becomes available.
 7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': Yes, each row is dedicated to a specific month and year, and how many homeless deaths occurred in that month.
+	- 'Toronto Shelter System Flow': Yes, each row is dedicated to a specific month, year and population group.
 8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
-	- TBD
+	- No.
 9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
-	- TBD
+	- No.
 10.
@@
 Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
-	- TBD
+	- Both datasets are self-contained.
 11.
@@
 If so, please provide a description.*
-	- TBD
+	- No.
 12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
-	- TBD
+	- No.
 13.
@@
 If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
-	- TBD
+	- 'Toronto Shelter System Flow': Yes, there are 6 population groups (Chronic, Refugees, Families, Youth, Single Adult, Non-Refugees) which are further divided into 14 categories (returned_from_housing, returned_to_shelter, age16-24, gender_male etc.)
 14.
@@
 If so, please describe how.*
-	- TBD
+	- No.
 15.
@@
 If so, please provide a description.*
-	- TBD
+	- No.
 16. *Any other comments?*
-	- TBD
+	- No.
 
@@
 If so, please describe how.*
-	- TBD
+	- Directly observable and collected from subjects (for homeless shelters through SMIS)
 2.
@@
 How were these mechanisms or procedures validated?*
-	- TBD
+  - 'Deaths of People Experiencing Homelessness': TPH collects, analyzes and reports data. SSHA share information about death with TPH.
+	- 'Toronto Shelter System Flow': Usage of SMIS software.
 3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
-	- TBD
+	- Not a sample from larger dataset.
 4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': deceased homeless people of Toronto. Not compensated.
+	- 'Toronto Shelter System Flow': Homeless Shelter Users of Toronto. Not compensated.
 5.
@@
 If not, please describe the timeframe in which the data associated with the instances was created.*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': January 2017 to June 2023.
+	- 'Toronto Shelter System Flow': January 2018 to February 2024.
 6.
@@
 If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- TBD
+	- No.
 7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
-	- TBD
+	- Obtained from Open Data Toronto.
 8.
@@
 If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
-	- TBD
+	- Data collection notification is subject to the policies of organizations collecting data.
 9.
@@
 If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
-	- TBD
+	- Data collection consent is subject to the policies of organizations collecting data.
 10.
@@
 If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
-	- TBD
+	- Data collection consent revocation is subject to the policies of organizations collecting data.
 11.
@@
 If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
-	- TBD
+	- No.
 12. *Any other comments?*
-	- TBD
+	- No.
 
@@
 If not, you may skip the remaining questions in this section.*
-	- TBD
+	- Not to my knowledge.
 2.
@@
 If so, please provide a link or other access point to the "raw" data.*
-	- TBD
+	- https://open.toronto.ca/dataset/deaths-of-people-experiencing-homelessness/ and https://open.toronto.ca/dataset/toronto-shelter-system-flow/
 3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
-	- TBD
+	- No.
 4. *Any other comments?*
-	- TBD
+	- No.
 
@@
 1. *Has the dataset been used for any tasks already? If so, please provide a description.*
-	- TBD
+	- No.
 2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
-	- TBD
+	- No.
 3. *What (other) tasks could the dataset be used for?*
-	- TBD
+	- Both datasets can be used to analyze homeless populations, and factors affecting their mortality, and shelter usage trends.
 4.
@@
 Is there anything a dataset consumer could do to mitigate these risks or harms?*
-	- TBD
+	- No.
 5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
-	- TBD
+	- No.
 6. *Any other comments?*
-	- TBD
+	- No.
 
@@
 If so, please provide a description.*
-	- TBD
+	- No.
 2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
-	- TBD
+	- Both datasets are available for public use through OpenDataToronto's portal.
 3. *When will the dataset be distributed?*
-	- TBD
+	- They are available to use currently.
 4.
@@
 If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
-	- TBD
+	- Both datasets are subject to Ontario Open Government License.
 5.
@@
 If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
-	- TBD
+	- No.
 6.
@@
 If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
-	- TBD
+	- No.
 7. *Any other comments?*
-	- TBD
+	- No.
 
@@
 1. *Who will be supporting/hosting/maintaining the dataset?*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': maintained by OpenDataToronto, TPH and SSHA.
+	- 'Toronto Shelter System Flow': maintained by OpenDataToronto, and SMIS.
 2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
-	- TBD
+  - 'Deaths of People Experiencing Homelessness': EDAU@toronto.ca
+	- 'Toronto Shelter System Flow': tsssdata@toronto.ca
 3. *Is there an erratum? If so, please provide a link or other access point.*
-	- TBD
+	- No.
 4.
@@
 If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
-	- TBD
+	- 'Deaths of People Experiencing Homelessness': updated by maintainers semi-annually on the OpenDataToronto portal.
+	- 'Toronto Shelter System Flow': updated by maintainers monthly on the OpenDataToronto portal.
 5.
@@
 If so, please describe these limits and explain how they will be enforced.*
-	- TBD
+	- No.
 6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
-	- TBD
+	- Both datasets have new data just appended upon previous instances.
 7.
@@
 If so, please provide a description.*
-	- TBD
+	- Both datasets are available to be downloaded in various formats for extension/augmentation purposes.
 8. *Any other comments?*
-	- TBD
+	- No.
 

### 2024-04-18 other/datasheet/datasheet.qmd
fix date

@@
 thanks: "Code and data are available at: https://github.com/Maroosh-Gillani/homeless_mortality"
-date: "`r Sys.time()`"
-date-format: "D MMMM YYYY"
+date: today
+date-format: long
 format: pdf
